---
title: "Data Processing (fastq to bams) for sequencing data produced by Sard and Homola"
author: "Nicole Adams"
date: '2022-06-21'
output: 
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE, echo=TRUE}
# Load necessary packages
library(tidyverse)
library(data.table)
library(viridisLite)
library(readxl)
library(lubridate)
library(kableExtra)
library(ggrepel)
library(ggpubr)

# Define Rmd knit options
knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(
	fig.height = 6,
	fig.width = 8,
	warning = FALSE, # suppress warnings
	message = FALSE # suppress messages
)
#echo = FALSE # defaults to not showing R code, change chunk to echo=TRUE where wanted

# Set eval=FALSE for all bash chunks
knitr::opts_hooks$set(
  engine = function(options) {
    if (options$engine == "bash") {
      options$eval <- FALSE
    }
    options
  }
)
```

# Seq processesing for 2022 sequencing run of Red Swamp Crayfish
Samples were sequenced at MSU's RTSF Genomics Core, project ID ADA12634 (RSC_novaSeq_2022). We submitted a single pool made up of 24 Illumina compatible libraries. The pool was QC’d and quantified using a combination of Qubit dsDNA HS, Agilent 4200 TapeStation HS DNA1000 and Invitrogen Collibri Library Quantification qPCR assays. The pool was loaded onto one (1) lane of a NovaSeq S4 flow cell and sequencing performed in a 2x150bp paired end read format using a NovaSeq 6000 v1.5 300 cycle reagent kit. Base calling was done by Illumina Real Time Analysis (RTA) v3.4.4 and output of RTA was demultiplexed and converted to FastQ format with Illumina Bcl2fastq v2.20.0.

&nbsp;

## Download raw sequence data from RTSF
Bash commands for getting raw data from MSU RTSF to MSU HPCC. Following the RTSF [instructions](https://rtsf.natsci.msu.edu/genomics/data-retrieval/#HPCC)
```{bash, eval=FALSE, message=FALSE}
# FTPS transfer from the RTSF FTP server to the HPCC only works if you are currently working from the rsync login node. Data for this run is in subdirectory 20220616_DNASeq_PE150

cd /mnt/home/adamsn23/RedSwampCrayfish/rawData2022

module load GCCcore/6.4.0 wget/1.19.4

wget -r -np -nH --ask-password ftps://scribnerk@titan.bch.msu.edu/20220616_DNASeq_PE150

# check files with md5
for FILE in *gz; do md5sum $FILE; done

# in excel I compared content of md5.txt with the result of md5 check and they were the same
```

&nbsp;

### Organize directories
```{bash, eval=FALSE, message=FALSE}
# /Users/neasci/Documents/crayfish_lab/RSCseq2022

# I renamed rawData2022 and the raw data directory
mv rawData2022/ seq2022/ 
mv 20220616_DNASeq_PE150/ rawData2022/ 
```

&nbsp;
&nbsp;
&nbsp;

## Pre-processing 

&nbsp;

### Make directories needed for flipReads
```{bash, eval=FALSE}
# Create directories needed for 2_flipReads_NEA.sh
mkdir /mnt/home/adamsn23/RedSwampCrayfish/seq2022/scripts
mkdir /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT
mkdir /mnt/home/adamsn23/RedSwampCrayfish/seq2022/SHELL
mkdir /mnt/home/adamsn23/RedSwampCrayfish/seq2022/SHELL/dependencies

#mkdir /mnt/gs18/scratch/users/adamsn23/RSC/
mkdir /mnt/gs21/scratch/adamsn23/RSC/seq2022 # new HPCC scratch space, used to be /mnt/gs21/scratch/gs18

```

&nbsp;
&nbsp;

## Unzip and flip reads
(.../RedSwampCrayfish/seq2022/scripts/2_flipReads_NEA.sh)
```{bash, eval=FALSE}
#==================================================================================================
#   File: Flip_reads.sh
#   Date: 01/05/21, 02/03/2022 (NEA), 06/28/2022 (NEA)
#   Description: Flip the orientation of the Best RAD reads
#--------------------------------------------------------------------------------------------------
#       Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
DATA_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022/rawData2022
my_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs21/scratch/adamsn23/RSC/seq2022   #changed from /mnt/gs18/scratch/users/adamsn23/RSC/seq2022 

mkdir $my_PATH/SHELL/flipReads
mkdir $my_PATH/QSTAT/flipReads

mkdir $SCR/unzipped/
mkdir $SCR/flipped/ #this line wasn't in Jared's script


### Unzip the files
cd $my_PATH/SHELL/flipReads

find $DATA_PATH -name '*.fastq.gz' | while read -r LINE
do
dat=$(echo $LINE | awk -F "/" '{print $8}' | awk -F ".fastq" '{print $1}')
datn=$(echo $LINE | awk -F "/" '{print $8}' | awk -F "_" '{print $2}') 

echo '#!/bin/sh 
#SBATCH -N 1
#SBATCH -t 1:00:00
#SBATCH --mem 64G 
#SBATCH -J '$datn'_unzip
#SBATCH -o '$my_PATH'/QSTAT/flipReads/'$datn'_unzip.o
#SBATCH --error='$my_PATH'/QSTAT/flipReads/'$datn'_unzip.err

## Unzip
gunzip -c '$LINE' > '$SCR'/unzipped/'$dat'.fastq

scontrol show job ${SLURM_JOB_ID} $' > "$datn"_unzip.sh

sbatch "$datn"_unzip.sh

done


#### Flip reads

cd $my_PATH/SHELL/flipReads

find $DATA_PATH -name '*1_001.fastq.gz' | while read -r LINE
do
dat=$(echo $LINE | awk -F "/" '{print $8}' | awk -F ".fastq" '{print $1}')
datn=$(echo $LINE | awk -F "/" '{print $8}' | awk -F "_" '{print $2}') 
num=$(echo $LINE | awk -F "/" '{print $8}' | awk -F "_" '{print $3}')

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 12:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=64G
#SBATCH -J '$datn'_flipReads
#SBATCH -o '$my_PATH'/QSTAT/flipReads/'$datn'_flipReads.o
#SBATCH --error='$my_PATH'/QSTAT/flipReads/'$datn'_flipReads.err

perl '$my_PATH'/SHELL/dependencies/bRAD_flip_trim.pl \
    '$my_PATH'/SHELL/dependencies/'$datn'_barcodes.txt \
    '$SCR'/unzipped/RSC_'$datn'_'$num'_L004_R1_001.fastq \
    '$SCR'/unzipped/RSC_'$datn'_'$num'_L004_R2_001.fastq \
    '$SCR'/flipped/'$datn'_flipped.1.fq \
    '$SCR'/flipped/'$datn'_flipped.2.fq
    

scontrol show job ${SLURM_JOB_ID} $' > "$datn"_flipReads.sh
sbatch "$datn"_flipReads.sh

done

```

&nbsp;

Run jobs
```{bash, eval=FALSE}
bash scripts/2_flipReads_NEA.sh
```

&nbsp;

### Zip the flipped reads
(.../RedSwampCrayfish/seq2022/scripts/3_zipFlipped_NEA.sh)
```{bash, eval=FALSE}
#==================================================================================================
#   File: Zip_flipped.sh
#   Date: 01/06/21, 02/07/2022 (NEA), 06/28/2022 (NEA)
#   Description: Zip the flipped best RAD reads
#--------------------------------------------------------------------------------------------------
#       Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
my_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

cd $my_PATH/SHELL/flipReads

ls $SCR/flipped | grep -v ".2.fq" | awk -F "_flipped" '{print $1}' | while read -r LINE
do

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 3:59:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=64G
#SBATCH -J Zip_'$LINE'
#SBATCH -o '$my_PATH'/QSTAT/flipReads/Zip_'$LINE'.o
#SBATCH --error '$my_PATH'/QSTAT/flipReads/Zip_'$LINE'.err

cd '$SCR'/flipped

gzip < ./'$LINE'_flipped.1.fq > '$LINE'_flipped.1.fq.gz
gzip < ./'$LINE'_flipped.2.fq > '$LINE'_flipped.2.fq.gz

scontrol show job ${SLURM_JOB_ID} $' > Zip_"$LINE".sh

sbatch Zip_"$LINE".sh

done

```

&nbsp;
&nbsp;

## Demultiplex the libraries into individual samples
(.../RedSwampCrayfish/seq2022/scripts/4_demultiplex_NEA.sh)
```{bash, eval=FALSE}
#==================================================================================================
#   File: 4_demultiplex.sh
#   Date: 01/06/21,  02/08/2022 (NEA), 06/28/2022 (NEA)
#   Description: Demultiplex libraries using process_radtags
#--------------------------------------------------------------------------------------------------
#   Author: Jared Homola, Nicole Adams
#==================================================================================================

RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs21/scratch/adamsn23/RSC/seq2022 # changed from /mnt/gs18/scratch/users/adamsn23/RSC/seq2022

mkdir $RUN_PATH/SHELL/demult
mkdir $RUN_PATH/QSTAT/demult
mkdir $SCR/demult

cd $RUN_PATH/SHELL/demult

### Demultiplex 
ls $SCR/flipped | grep -v ".gz" | grep -v ".2.fq" | awk -F "_flipped" '{print $1}' | while read -r LINE
do

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 5:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=24G
#SBATCH -J '$LINE'_demultiplex
#SBATCH -o '$RUN_PATH'/QSTAT/demult/'$LINE'_demultiplex.o
#SBATCH --error '$RUN_PATH'/QSTAT/demult/'$LINE'_demultiplex.err

module purge
module load GCC/9.3.0 OpenMPI/4.0.3 Stacks/2.59

process_radtags \
    -1 '$SCR'/flipped/'$LINE'_flipped.1.fq.gz \
    -2 '$SCR'/flipped/'$LINE'_flipped.2.fq.gz \
    -i gzfastq \
    -y gzfastq \
    -o '$SCR'/demult/ \
    -b '$RUN_PATH'/SHELL/dependencies/'$LINE'_barcodes.txt \
    --inline_null \
    -e sbfI \
    --barcode_dist_1 1 \
    --retain_header

scontrol show job ${SLURM_JOB_ID}' > demultiplex_"$LINE".sh

sbatch demultiplex_"$LINE".sh

done
```

&nbsp;

### Combine demultiplex results
```{bash, eval=FALSE}
cd /mnt/home/adamsn23/RedSwampCrayfish/seq2022/

### Summarize demult results

echo "LIB totSequences retainedReads percRetained" > OUT/demultRes.txt

for FILE in QSTAT/demult/Lib*_demultiplex.err
do
ind=$(echo $FILE | awk -F "/" '{print $3}' | awk -F "_" '{print $1}')
res1=$(grep 'total sequences' $FILE | grep -Eo '[0-9]{1,}' | tr -s '\n' '\t')
res2=$(grep 'retained reads (' $FILE | grep -Eo '[0-9]{1,}' | tr -s '\n' '\t')
echo $ind $res1 $res2 >> OUT/demultRes.txt
done

```

&nbsp;
&nbsp;

## Remove PCR clones
(.../RedSwampCrayfish/seq2022/scripts/5_removeClones_NEA.sh)
Since the number of samples exceeds how many jobs that can be run on the HPCC I split them up: grep "group1", grep "group2A", grep "group2B"; then did grep -v "group1" | grep -v "group2A" | grep -v "group2B" to get the rest
```{bash, eval=FALSE}
#==================================================================================================
#   File: removeClones.sh
#   Date: 01/06/21, 02/08/2022 (NEA), 06/28/2022 (NEA)
#   Description: Filter clonal reads from data
#--------------------------------------------------------------------------------------------------
#	Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

mkdir $SCR/cloneFilter
mkdir $RUN_PATH/SHELL/cloneFilter
mkdir $RUN_PATH/QSTAT/cloneFilter
mkdir $RUN_PATH/OUT

cd $RUN_PATH/SHELL/cloneFilter

# too many files to submit to slurm so did FC and XX separate, then did the rest
ls $SCR/demult/ | grep "fq" | grep -v ".2.fq.gz" | grep -v "rem" | grep "FC" | sed 's/\.1\.fq\.gz//g' | sort | uniq | while read -r LINE
#ls $SCR/demult/ | grep "fq" | grep -v ".2.fq.gz" | grep -v "rem" | grep -v "PC" | grep -v "FC" | sed 's/\.1\.fq\.gz//g' | sort | uniq | while read -r LINE
do

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 1:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=32G
#SBATCH -J '$LINE'_cloneFilter
#SBATCH -o '$RUN_PATH'/QSTAT/cloneFilter/cloneFilter.'$LINE'.o
#SBATCH --error '$RUN_PATH'/QSTAT/cloneFilter/cloneFilter.'$LINE'.err

module purge
module load GCC/9.3.0 OpenMPI/4.0.3 Stacks/2.59

clone_filter -1 '$SCR'/demult/'$LINE'.1.fq.gz \
    -2 '$SCR'/demult/'$LINE'.2.fq.gz \
    -i gzfastq \
    -o '$SCR'/cloneFilter/

scontrol show job ${SLURM_JOB_ID}' > ./cloneFilter."$LINE".sh

sbatch ./cloneFilter."$LINE".sh

done


### Summarize results
##grep '% clone reads' $RUN_PATH/QSTAT/cloneFilter/cloneFilter*.err >> $RUN_PATH/OUT/cloneFilterRes.txt

#for FILE in QSTAT/cloneFilter/cloneFilter*.err 
#do 
#ind=$(echo $FILE | awk -F "/" '{print $3}' | awk -F "." '{print $2}')  
##perc=$(grep -oP '.{0,5}%' $FILE) 
#res=$(grep '% clone'  $FILE | grep -Eo '[0-9]{1,}' | tr -s '\n' '\t') 
#echo $ind $res >> OUT/cloneFilterRes.txt 
#done 
```

&nbsp;

### Look at clone results
```{r}
cl <- read.delim("~/Documents/crayfish_lab/RSCseq2022/cloneFilterRes.txt", header = F) #N=2284

cl2 <- cl %>% separate(V1, into = c("ind", "inputPairs", "outputPairs", "discardedPairs", "clone1", "clone2"), sep = " ") %>% unite("percCloneReads", clone1:clone2, sep = ".") 

cols2chg <- c("inputPairs", "outputPairs", "discardedPairs", "percCloneReads")
cl2[cols2chg] <- sapply(cl2[cols2chg],as.numeric)

meta.df2 <- cl2

# Plot distribution of percent clone reads
cl.qts <- quantile(meta.df2$percCloneReads,probs=c(.05,.95), na.rm = T)

percClone.p <- ggplot(meta.df2, aes(percCloneReads)) +
    geom_histogram(binwidth = 2) +
  geom_vline(xintercept = cl.qts[1], color="red", linetype="dashed") +
  geom_vline(xintercept = cl.qts[2], color="red", linetype="dashed") +
    theme_minimal()

# Plot percent clone reads per individual
percClone.p2 <- ggplot(meta.df2, aes(x=reorder(ind, -percCloneReads), y=percCloneReads)) +
    geom_point() +
    theme_minimal() +
  labs(x="Individual") +
  theme(axis.text.x = element_blank(), panel.grid.major = element_blank())


# Summarize percent clone reads per individual
cl.tabA <- meta.df2 %>% summarise(mean=mean(percCloneReads), sd=sd(percCloneReads)) # 69%

# list samples with high number of clones
high.clone <- meta.df2 %>% filter(percCloneReads > cl.qts[2])


high.clone %>% dplyr::select(ind, percCloneReads) %>% kbl(
  caption = "Samples with high amount of clone reads",
  col.names = c("Individual ID", "Percent clone reads"), # Clean column names
  align = 'c',
  format = "html")  %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), # Popular styling options
    full_width = T)

ggarrange(percClone.p, percClone.p2)

```

&nbsp;
&nbsp;

## Trim and filter
(.../RedSwampCrayfish/seq2022/scripts/6_trimQualityFilter_NEA.sh)
Since the number of samples exceeds how many jobs that can be run on the HPCC I split them up: grep "FC", grep "MB", grep "SHD"; then did grep -v "FC" | grep -v "MB" | grep -v "SHD" to get the rest
```{bash, eval=FALSE}
#==================================================================================================
#   File: 6_trimQualityFilter.sh
#   Date: 01/07/20, 02/10/2022 (NEA), 06/29/2022 (NEA)
#   Description: Use Trimmomatic to trim reads and quality filter
#--------------------------------------------------------------------------------------------------
#	Authors: Jared Homola, Nicole Adams
#==================================================================================================

RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

### Need to rename files by removing the extra .1 or .2 that was inserted
find $SCR/cloneFilter -type f -name '*.1.1.fq.gz' | while read FILE ; do
    newfile="$(echo ${FILE} | sed -e 's/\.1.fq.gz/.fq.gz/')" ;
    mv "${FILE}" "${newfile}" ;
done 

find $SCR/cloneFilter -type f -name '*.2.2.fq.gz' | while read FILE ; do
    newfile="$(echo ${FILE} | sed -e 's/\.2.fq.gz/.fq.gz/')" ;
    mv "${FILE}" "${newfile}" ;
done 


mkdir $SCR/trimmed
mkdir $RUN_PATH/SHELL/trimQualityFilter
mkdir $RUN_PATH/QSTAT/trimQualityFilter

cd $RUN_PATH/SHELL/trimQualityFilter

ls $SCR/cloneFilter | grep "fq" | grep -v ".2.fq.gz" | grep "FC" | sed 's/\.1\.fq\.gz//g' | while read -r LINE
do
echo '#!/bin/sh 
#SBATCH --nodes=1-4
#SBATCH --ntasks=1
#SBATCH -t 3:00:00
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8G
#SBATCH -J '$LINE'.trimQualityFilter
#SBATCH -o '$RUN_PATH'/QSTAT/trimQualityFilter/trimQualityFilter.'$LINE'.o
#SBATCH --error '$RUN_PATH'/QSTAT/trimQualityFilter/trimQualityFilter.'$LINE'.err

module purge
module load Trimmomatic/0.39-Java-11 Java/1.8.0_162

cd '$SCR'/cloneFilter
java -jar $EBROOTTRIMMOMATIC/trimmomatic-0.39.jar PE -threads 4 -phred33 \
'$LINE'.1.fq.gz '$LINE'.2.fq.gz \
'$SCR'/trimmed/'$LINE'_paired.1.fq.gz '$SCR'/trimmed/'$LINE'_unpaired.1.fq.gz \ '$SCR'/trimmed/'$LINE'_paired.2.fq.gz '$SCR'/trimmed/'$LINE'_unpaired.2.fq.gz \
ILLUMINACLIP:/mnt/research/Scribner_Lab/projects/RedSwampCrayfish_MISGP/SHELL/dependencies/adapters.fa:2:30:10 SLIDINGWINDOW:4:15 MINLEN:50 

scontrol show job ${SLURM_JOB_ID}' > ./trimQualityFilter."$LINE".sh

sbatch ./trimQualityFilter."$LINE".sh

done


###### How many reads were trimmed and filtered? ######
#grep 'Surviving' $RUN_PATH/QSTAT/trimQualityFilter/trimQualityFilter*.o >> $RUN_PATH/OUT/trimQualityFilterRes.txt

#for FILE in QSTAT/trimQualityFilter/trimQualityFilter*.err
#do
#ind=$(echo $FILE | awk -F "/" '{print $3}' | awk -F "." '{print $2}')
#res=$(grep 'Surviving'  $FILE | awk -F " " '{print $4 "\t" $7 "\t" $8 "\t" $12 "\t" $13 "\t" $17 "\t" $18 "\t" $20 "\t" $21}')
#echo $ind $res >> OUT/trimQualityFilterRes.txt
#done

```

&nbsp;
&nbsp;

## Mapping to reference genome
### Make files needed for mapping
```{bash, eval=FALSE}
mkdir $SCRATCH/RSC/seq2022/trimmed/paired

mv $SCRATCH/RSC/seq2022/trimmed/*_paired* $SCRATCH/RSC/seq2022/trimmed/paired

```

&nbsp;

### Map
(.../RedSwampCrayfish/seq2022/scripts/8_bwaMemMap_NEA.sh)
```{bash, eval=FALSE}
#==================================================================================================
#   File: bwaMemMap.sh
#   Date: 01/07/21, 02/11/2022 (NEA), 07/04/2022 (NEA)
#   Description: Map reads to reference genome using bwa-mem
#--------------------------------------------------------------------------------------------------
#	Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

mkdir $SCR/mapped
mkdir $RUN_PATH/SHELL/mapping
mkdir $RUN_PATH/QSTAT/mapping

cd $RUN_PATH/SHELL/mapping
ls $SCR/trimmed/paired | grep "fq" | grep -v ".2.fq.gz" | sed 's/\_paired.1\.fq\.gz//g' | grep "FC" | uniq | while read -r LINE
do

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 3:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=36G
#SBATCH -J '$LINE'.bwaMemMap
#SBATCH -o '$RUN_PATH'/QSTAT/mapping/'$LINE'.bwaMemMap.o
#SBATCH --error '$RUN_PATH'/QSTAT/mapping/'$LINE'.bwaMemMap.err

module purge
module load GCC/6.4.0-2.28 OpenMPI/2.1.1 BWA/0.7.17 SAMtools/1.9

cd '$SCR'/trimmed/paired

bwa mem -R "@RG\tID:'$LINE'\tSM:'$LINE'\tPL:ILLUMINA\tLB:LB1" /mnt/home/adamsn23/RedSwampCrayfish/reference/GCF_020424385.1_ASM2042438v2_genomic.fna ./'$LINE'_paired.1.fq.gz ./'$LINE'_paired.2.fq.gz | samtools view -Sb - > ../../mapped/'$LINE'.pe.bam 

scontrol show job ${SLURM_JOB_ID}' > ./bwaMemMap."$LINE".sh

sbatch ./bwaMemMap."$LINE".sh

done

```

&nbsp;

### Sort mapped BAMs
(.../RedSwampCrayfish/seq2022/scripts/10_sortBAM_NEA.sh)
```{bash, eval=FALSE}
#==================================================================================================
#   File: 10_sortBAM.sh
#   Date: 01/07/20, 02/15/2022 (NEA), 07/04/2022 (NEA)
#   Description: Sort BAM files
#--------------------------------------------------------------------------------------------------
#	Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

mkdir $RUN_PATH/QSTAT/sortBAM
mkdir $RUN_PATH/SHELL/sortBAM_jobs
mkdir $RUN_PATH/OUT/sorted

cd $RUN_PATH/SHELL/sortBAM_jobs
ls $SCR/mapped | grep bam | awk -F "." '{print $1}' | grep "SHD" | sort | uniq | while read -r LINE
do
echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 1:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=8G
#SBATCH -J '$LINE'.sortBAM
#SBATCH -o '$RUN_PATH'/QSTAT/sortBAM/'$LINE'.sortBAM.o
#SBATCH --error '$RUN_PATH'/QSTAT/sortBAM/'$LINE'.sortBAM.err

module purge
module load GCC/9.3.0 SAMtools/1.15

samtools sort '$SCR'/mapped/'$LINE'.pe.bam -o '$RUN_PATH'/OUT/sorted/'$LINE'.pe.sorted.bam

scontrol show job ${SLURM_JOB_ID}' > ./sortBAM."$LINE".sh

sbatch ./sortBAM."$LINE".sh

done


```

&nbsp;

## Filter mapped BAMs
(.../RedSwampCrayfish/seq2022/scripts/11_filterBAMS_NEA.sh)
-q 20=keep quality >= 20; -h=keep header; -f2=keep proper pairs(PROPER_PAIR)? ; -F2308=exclude unmapped reads, not primary alignment, supplementary alignment.  see [this calculater](https://broadinstitute.github.io/picard/explain-flags.html)
```{bash, eval=FALSE}
#==================================================================================================
#   File: filterBAM.sh
#   Date: 01/07/21, 03/03/2022 (NEA), 07/05/2022 (NEA)
#   Description: Filter BAM files
#--------------------------------------------------------------------------------------------------
#	Authors: Jared Homola, Nicole Adams
#==================================================================================================

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

mkdir $RUN_PATH/QSTAT/filterBAM
mkdir $RUN_PATH/SHELL/filterBAM_jobs
mkdir $RUN_PATH/OUT/filtered

cd $RUN_PATH/SHELL/filterBAM_jobs
ls $RUN_PATH/OUT/sorted | grep bam | awk -F "." '{print $1}' | grep "SHD" | sort | uniq | while read -r LINE
do

echo '#!/bin/sh 
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH -t 1:00:00
#SBATCH --cpus-per-task=1
#SBATCH --mem-per-cpu=8G
#SBATCH -J '$LINE'.filterBAM
#SBATCH -o '$RUN_PATH'/QSTAT/filterBAM/'$LINE'.filterBAM.o
#SBATCH --error '$RUN_PATH'/QSTAT/filterBAM/'$LINE'.filterBAM.err

module purge
module load GCC/9.3.0 SAMtools/1.15

cd '$RUN_PATH'/OUT/sorted
samtools view -q 20 -h -f2 -F2308  '$LINE'.pe.sorted.bam | samtools view -Sb > ../filtered/'$LINE'.pe.sorted.filtered.bam

scontrol show job ${SLURM_JOB_ID}' > ./filterBAM."$LINE".sh

sbatch ./filterBAM."$LINE".sh

done
```

&nbsp;

## Mapping metrics
(.../RedSwampCrayfish/seq2022/scripts/11B_mapMetrics_NEA.sh)  
```{bash, eval=FALSE}
#!/bin/sh 
#SBATCH --nodes=1-4
#SBATCH --ntasks=6
#SBATCH -t 3:00:00
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8G
#SBATCH -J mapMetrics
#SBATCH -o /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/mapMetrics.o
#SBATCH --error /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/mapMetrics.err

#==================================================================================================
#   File: 11B_mapMetrics_NEA.sh
#   Date: 03/22/2022, 07/05/2022
#   Description: Get summary of mapping
#--------------------------------------------------------------------------------------------------
#       Authors: Nicole Adams
#==================================================================================================

module purge
module load GCC/9.3.0 SAMtools/1.15

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022
BAM_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022/OUT/sorted


cd $RUN_PATH/SHELL

touch $BAM_PATH/mapSummary_seq2022.txt

for sample in `ls $BAM_PATH/*sorted.bam | cut -f1 -d'.'`
do
  samtools flagstat "$sample".pe.sorted.bam -@ 6  > "$sample"_mapSum.txt
 awk 'FNR == 1{ print FILENAME }' "$sample"_mapSum.txt >> $BAM_PATH/mapSummary_seq2022.txt
 cat "$sample"_mapSum.txt >> $BAM_PATH/mapSummary_seq2022.txt

done

for sample in $BAM_PATH/*mapSum.txt; do awk 'FNR == 1{ print FILENAME } {printf "%-20s %-40s\n", $1, $3}' OFS="\t" $sample | awk '
{
    for (i=1; i<=NF; i++)  {
        a[NR,i] = $i
    }
}
NF>p { p = NF }
END {
    for(j=1; j<=p; j++) {
        str=a[1,j]
        for(i=2; i<=NR; i++){
            str=str" "a[i,j];
        }
        print str
    }
}' >> $BAM_PATH/mapSummary_seq2022.2.txt; done

grep 'mnt' $BAM_PATH/mapSummary_seq2022.2.txt > $BAM_PATH/mapSummary_seq2022.3.txt

rm $BAM_PATH/*_mapSum.txt
```

&nbsp;

### Repeat for filtered BAMs (11B_mapMetrics_filtered_NEA.sh)
```{bash, eval=FALSE}
#!/bin/sh 
#SBATCH --nodes=1-4
#SBATCH --ntasks=8
#SBATCH -t 3:00:00
#SBATCH --cpus-per-task=6
#SBATCH --mem-per-cpu=8G
#SBATCH -J mapMetrics
#SBATCH -o /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/mapMetrics.o
#SBATCH --error /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/mapMetrics.err

#==================================================================================================
#   File: 11B_mapMetrics_NEA.sh
#   Date: 03/22/2022, 07/05/2022
#   Description: Get summary of mapping
#--------------------------------------------------------------------------------------------------
#       Authors: Nicole Adams
#==================================================================================================

module purge
module load GCC/9.3.0 SAMtools/1.15

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022
BAM_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022/OUT/filtered


cd $RUN_PATH/SHELL

touch $BAM_PATH/mapSummary_seq2022_filt.txt

for sample in `ls $BAM_PATH/*filtered.bam | cut -f1 -d'.'`
do
  samtools flagstat "$sample".pe.sorted.filtered.bam -@ 6  > "$sample"_mapSum.txt
 awk 'FNR == 1{ print FILENAME }' "$sample"_mapSum.txt >> $BAM_PATH/mapSummary_seq2022_filt.txt
 cat "$sample"_mapSum.txt >> $BAM_PATH/mapSummary_seq2022_filt.txt

done

for sample in $BAM_PATH/*mapSum.txt; do awk 'FNR == 1{ print FILENAME } {printf "%-20s %-40s\n", $1, $3}' OFS="\t" $sample | awk '
{
    for (i=1; i<=NF; i++)  {
        a[NR,i] = $i
  }
}
NF>p { p = NF }
END {
    for(j=1; j<=p; j++) {
        str=a[1,j]
        for(i=2; i<=NR; i++){
            str=str" "a[i,j];
        }
        print str
    }
}' >> $BAM_PATH/mapSummary_seq2022_filt.2.txt; done

grep 'mnt' $BAM_PATH/mapSummary_seq2022_filt.2.txt > $BAM_PATH/mapSummary_seq2022_filt.3.txt

rm $BAM_PATH/*_mapSum.txt

```

&nbsp;

#### Examine mapping results in R
I ran summary stats for sorted bams (.pe.sorted.bam) and filtered bams (.pe.sorted.filtered.bam)
```{r}
# Read in mapping summary data
msumA <- read.table("~/Documents/crayfish_lab/RSCseq2022/mapSummary_seq2022.3.txt")

# Clean up sample names
colnames(msumA) <- c("Sample", "QCpassedReads", "primary", "secondary", "supplementary", "duplicates", "primaryDuplicates", "mapped", "primaryMapped", "paired", "read1", "read2", "properlyPaired", "itselfYmateMapped", "singletons", "mateMappedDiffChr", "mateMappedDiffChr_mapQ5")

nothanks <- c("a", "b", "c", "d", "e", "f", "g", "h")
msum <- msumA %>% separate(Sample, c("a", "b", "c", "d", "e", "f", "g", "h", "i"), sep = "/") %>% select(-nothanks) %>%
  separate(i, c("Sample", "stuff"), sep = "_") %>% select(-stuff)

# Calculate percentages
msum$percentMap <- (msum$mapped/msum$QCpassedReads)*100
msum$percentPaired <- (msum$properlyPaired/msum$paired)*100
msum$percentSingle <- (msum$singletons/msum$properlyPaired)*100


# Assign species labels
other_sp <- c("Cfod", "Ctho", "Fprop", "Fvir", "Lpol", "Pacu")
pattern <- paste(other_sp, collapse = "|")

msum <- msum %>%
  mutate(
    species = if_else(
      str_detect(Sample, pattern),
      "OtherSpecies",
      "Pclarkii"
    )
  ) %>%
  mutate(
    species = factor(species, levels = c("Pclarkii", "OtherSpecies"))) %>%
  filter(!Sample == "empty")

# Summary by species (P. clarkii vs other species)
msum_species_summary <- msum %>%
  group_by(species) %>%
  summarise(
    n_samples = n(),
    meanPercentMap = mean(percentMap, na.rm = TRUE),
    sdPercentMap = sd(percentMap, na.rm = TRUE),
    meanPercentPaired = mean(percentPaired, na.rm = TRUE),
    sdPercentPaired = sd(percentPaired, na.rm = TRUE),
    .groups = "drop"
  )


kable(msum_species_summary, digits = 2, 
      caption = "Mapping summary: P. clarkii vs Other Species") %>% 
  kable_styling()


# Create histogram
msum.p <- ggplot(msum, aes(x = percentMap, fill = species)) +
  geom_histogram(color = "gray40", alpha = 0.7, position = "identity", bins = 30) +
  facet_wrap(~species, ncol = 1, scales = "free_y") +
  labs(
    title = "Distribution of Mapping Percentages",
    x = "Percent Mapped",
    y = "Count",
    fill = "Species"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

msum.p
```

&nbsp;
&nbsp;

### Read counts
(.../RedSwampCrayfish/seq2022/scripts/12_countReads_NEA.sh)
```{bash, eval=FALSE}
#!/bin/sh 
#SBATCH --nodes=1-4
#SBATCH --ntasks=8
#SBATCH -t 1:00:00
#SBATCH --cpus-per-task=4
#SBATCH --mem-per-cpu=8G
#SBATCH -J Count_reads
#SBATCH -o /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/Count_reads.o
#SBATCH --error /mnt/home/adamsn23/RedSwampCrayfish/seq2022/QSTAT/Count_reads.err

#==================================================================================================
#   File: Count_reads.sh
#   Date: 04/22/19, 03/11/2022 (NEA), 07/05/2022 (NEA)
#   Description: Flip the orientation of the Best RAD reads
#       Run: Interactively - array
#--------------------------------------------------------------------------------------------------
#       Authors: Jared Homola, Seth Smith, Nicole Adams
#==================================================================================================

#Define alias for project root directory
RUN_PATH=/mnt/home/adamsn23/RedSwampCrayfish/seq2022
SCR=/mnt/gs18/scratch/users/adamsn23/RSC/seq2022

cd $RUN_PATH/SHELL

touch $RUN_PATH/OUT/Read_counts.txt

# old way
#ls $SCR/RSC/trimmed/paired | grep "fq" | grep -v ".2.fq.gz" | grep -v "rem" | awk -F "." '{print $1}' | sort | uniq | while read -r LINE
#do
#   echo $LINE >> $RUN_PATH/OUT/Read_counts.txt
#   zcat $SCR/RSC/trimmed/paired/$LINE.1.fq.gz | echo $((`wc -l`/4)) >> #$RUN_PATH/OUT/Read_counts.txt
#done 

# new way, puts name and number on same line
ls $SCR/trimmed/paired | grep "fq" | grep -v ".2.fq.gz" | grep -v "rem" | awk -F "." '{print $1}' | sort | uniq | while read -r LINE 
do 
ind=$(echo $LINE) 
dat=$(zcat $SCR/trimmed/paired/$LINE.1.fq.gz | echo $((`wc -l`/4))) 
echo $ind $dat >> $RUN_PATH/OUT/Read_counts.txt 
done  

```

&nbsp;

### Examine read counts in R
```{r}
counts.a <- as.data.frame(read.delim("~/Documents/crayfish_lab/RSCseq2022/Read_counts.txt", header = FALSE))

counts.b <- counts.a %>% separate(V1, into = c("ind", "reads"), sep = " ") %>%
  separate(ind, into = c("ind", "stuff"), sep = "_") %>% select(-stuff)

# some samples are present twice in the file... so remove them
counts.c <- counts.b %>% distinct(ind, .keep_all = T)

# add in pop names and remove empty
counts <- counts.c %>% mutate(pop = ind) %>% separate(pop, into = c("pop", "extra"), sep = "-") %>% select(-extra) %>% filter(!ind == "empty") %>% mutate(reads= as.numeric(reads))


# identify low read samples:
qts <- quantile(counts$reads,probs=c(.05,.95))

counts.p <- ggplot(counts, aes(x=reads)) +
  geom_histogram() +
  geom_vline(xintercept = qts[1], color="red", linetype="dashed") +
  geom_vline(xintercept = qts[2], color="red", linetype="dashed") +
  theme_minimal() +
  theme(text = element_text(size = 16)) 

# identify low read samples:
low.counts <- counts %>% filter(reads < qts[1]) %>% arrange(reads)

no_low <- dim(low.counts)[1]

low.counts %>% dplyr::select(ind, reads) %>% kbl(
  caption = paste0("Samples with Low Read Counts (N=", no_low, ")"), # Add a title
  col.names = c("Individual ID", "Read Count"), # Clean column names
  align = 'c',
  format = "html")  %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"), # Popular styling options
    full_width = T)
```