---
title: "PCA and DAPC analyses"
author: "Nicole Adams"
date: "2025-04-05"
output: 
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE, echo=TRUE}
# Load necessary packages
library(tidyverse)
library(reshape2)
library(ggrepel)
library(viridis)
library(RColorBrewer)
library(ggpubr)
#BiocManager::install("SeqArray")
library(SeqArray)
library(SNPRelate)
library(vcfR)
library(adegenet)

# Set working directory
knitr::opts_knit$set(root.dir = "~/Documents/crayfish_lab/RedSwampCrayfish_geneFlow/")

# Define Rmd knit options
#knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(
	fig.height = 6,
	fig.width = 8,
	warning = FALSE, # suppress warnings
	message = FALSE # suppress messages
)
#echo = FALSE # defaults to not showing R code, change chunk to echo=TRUE where wanted

# Set eval=FALSE for all bash chunks
knitr::opts_hooks$set(
  engine = function(options) {
    if (options$engine == "bash") {
      options$eval = FALSE
      message = FALSE
    }
    options
  }
)
```

```{r setup, include=FALSE, echo=TRUE}
# Source functions to fix population names
source("sensitive/population_name_functions.R")
```


# Conduct PCA and DAPC analyses on RSC samples
Code to conduct PCA and DAPC analyses on RSC samples sequenced by Homola and Adams. For previous data processing and genotyping steps see dataProcessing_fastq2bams and genotyping_filtering files.<br>

# Run PCA in PLINK
```{bash}
module load PLINK/2.00a_10.2-x86_64

mkdir plinkAnalyses
 
# generate plink files 
plink2 --vcf jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb.recode.vcf --make-bed --allow-extra-chr --out plinkAnalyses/jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb

cd plinkAnalyses

# generate eigen values
plink2 --bfile jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb --pca --allow-extra-chr --out jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb.pca


# output: file.eigenvec and file.eigenval. The eigenval tells you in order of each PC ( so PC1,PC2….) the percentage each eigenvalue contributes to the variance. The eigenvec contains the coordinates for each sample. This file has no headers, and is tab seperated and contains the sample name in columns one and two, and then subsequently the eigenvals for each PC. 
```

&nbsp;

plot PCA
```{r plot PCA}
pca1x <- read.table("data/pca_dapc/jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb.pca.eigenval",sep=" ",header=F)

hist(pca1x$V1)
varexx <- pca1x %>% mutate(varex=V1/sum(V1), percVarex=round(V1/sum(V1)*100, 2))

pca2x <- read.table("data/pca_dapc/jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb.pca.eigenvec",sep="\t",header=F)


# fix names
pca2x$V2 <- gsub("FC815-212", "FC8-15-212",  pca2x$V2)

# add pops
pca2x.meta <- pca2x %>% mutate("sample"=V2) %>% separate(sample, into = c("pop", "stuff")) %>% dplyr::select(-stuff)

# fix more names and re-factor
pca2x.meta$pop <- convert_pop_names(pca2x.meta$pop)

# pca colored by indiv pop
pop_levels <- c("LOA", "EastGC11", "EastGC10", "EastGC7", "EastGC9", "EastGC2", "EastGC1", "EastGC3", "EastGC4", "EastGC13", "WestGC1", "WestGC2", "WestGC3", "Hotel5", "WestGC6", "Hotel1", "Hotel2", "Hotel3", "Hotel4", "Apt1", "Apt3", "Apt2", "Private1", "Drain1")

pca2x.meta$pop <- factor(pca2x.meta$pop, pop_levels)

# pca color by genetic group
pca2x.meta$deme <- case_when(
  pca2x.meta$pop %in% c("Drain1") ~ "Group5",
  pca2x.meta$pop %in% c("Apt1", "Apt2", "Apt3") ~ "Group1",
  pca2x.meta$pop %in% c("WestGC1", "WestGC2", "WestGC3", "WestGC6", "Hotel5") ~ "Group2_WestGC",
  pca2x.meta$pop %in% c("Hotel1", "Hotel2", "Hotel3", "Hotel4") ~ "Group2_Hotel",
  pca2x.meta$pop %in% c("EastGC11", "EastGC10", "EastGC7", "EastGC13", "EastGC1", "EastGC9") ~ "Group3_EastGC_SW",
  pca2x.meta$pop %in% c("EastGC2","EastGC3", "EastGC4") ~ "Group3_EastGC_NE",
  pca2x.meta$pop %in% c("Private1") ~ "Group4",
  TRUE ~ NA_character_
)

pca2x.meta$deme <- factor(pca2x.meta$deme, levels = c( "Group1", "Group2_WestGC", "Group2_Hotel", "Group3_EastGC_SW",  "Group3_EastGC_NE",  "Group4", "Group5"))

myCol <- brewer.pal(n = 10, name = 'Paired') 

pcaCols <- c("Group1" = myCol[8], "Group2_WestGC" = myCol[1], "Group2_Hotel" = myCol[2], "Group3_EastGC_SW" = myCol[3], "Group3_EastGC_NE" = myCol[4], "Group4" = myCol[6], "Group5" = myCol[9])
  

pclrk.pca.p3 <- ggplot(data=pca2x.meta, aes(V3,V4, color=deme)) +
  geom_point(size=5, alpha=0.75) +
  xlab(paste("PC1","(",varexx$percVarex[1],"%",")")) + ylab(paste("PC2","(",varexx$percVarex[2],"%",")")) +
  #scale_color_viridis_d() +
  scale_color_manual(values=pcaCols) +
  labs(color="Group") +
  theme_minimal() +
  theme(text = element_text(size=18)) 

pclrk.pca.p4 <- ggarrange(pclrk.pca.p3, labels="AUTO", font.label = list(size=20))

```

&nbsp;
&nbsp;

# Cluster analysis - DAPC
Read in VCF
```{r}
dat.vcf <-  read.vcfR(file="data/jy22_adults_baseFilter_imiss99_miss45.imiss70_hz.ab_1pRt_100kb.recode.vcf") 

dat.genind <- vcfR2genind(dat.vcf, sep = "[|/]") ## Create genind object

# get pop names
dat.pops <- dat.genind@tab[,1] %>%
  names() %>%
  as_tibble %>%
  separate(value, into = (c("Pop", NA))) 

# fix names, refactor, and reassign
rownames(dat.genind$tab) <- gsub("FC815-212", "FC8-15-212", rownames(dat.genind$tab))

dat.pops$Pop <- convert_pop_names(dat.pops$Pop)
dat.pops <- dat.pops %>% mutate(Pop = as.factor(Pop))

dat.genind@pop <- dat.pops$Pop
```

&nbsp;

```{r Overall DAPC}
# Run find.clusters - make sure MB2a is MB2
set.seed(21)

grp.stats <- find.clusters(dat.genind, max.n.clust=40, n.pca=200, choose.n.clust=FALSE)
grp <- find.clusters(dat.genind, max.n.clust=40, n.pca = 400, n.clust = 7)
# chose n.pca = 400, n.clusters= 7

grp.K <- grp.stats$Kstat %>% as.data.frame() %>% mutate(K=1:40) %>% rename("BIC"=".") %>%
ggplot(aes(x=K, y=BIC)) +
  geom_line(color="gray") +
  geom_point(shape=21, fill= "gray", size=3) +
  geom_vline(xintercept = length(grp$size), linetype="dashed", color="black") +
  #ggtitle("All sites") +
  theme_minimal() +
  theme(text=element_text(size=16))

grp.K.2 <- ggarrange(grp.K, labels = "C")

# dapc1 <- dapc.genind(x = dat.genind, pop = grp$grp, n.pca = 200, n.da = 6) # n.da=6
# 
# # Run optim a to find how many PCs to retain
# temp <- optim.a.score(dapc1)
# 
# # Rerun dapc with optim pcs
# dapc2 <- dapc.genind(x = dat.genind, pop = grp$grp, n.pca = 25, n.da = 6) # n.da=6

# Load dapc2 RDS
dapc2 <- readRDS("data/pca_dapc/dapcAll_dapc2.rds")

myCol <- brewer.pal(n = 10, name = 'Paired')
#display.brewer.pal(n = 10, name = 'Paired')
#scales::show_col(myCol)
#clusCols <- c("1" = myCol[2], "2" = myCol[6], "3" = myCol[4], "4" = myCol[1], "5" = myCol[7], "6" = myCol[10], "7" = myCol[8], "8" = myCol[3], "9" = myCol[9])
clusCols <- c("1" = myCol[2], "2" = myCol[9], "3" = myCol[6], "4" = myCol[3], "5" = myCol[8], "6" = myCol[1], "7" = myCol[4])
  

DAPC_2_probs <- round(dapc2$posterior, 2) %>%
  as_tibble(rownames = "SampleID") %>% 
  mutate(WaterbodyName = dat.genind@pop) %>% 
  pivot_longer(cols = 2:8,   ## change depending on the number of groups
               names_to = "Cluster", 
               values_to = "Probability")

DAPC_2_probs <- DAPC_2_probs %>%  mutate(WaterbodyName = fct_relevel(WaterbodyName, "Apt1", "Apt2", "Apt3", "WestGC1", "WestGC2", "WestGC3", "WestGC6", "Hotel5", "Hotel1", "Hotel2", "Hotel3", "Hotel4", "EastGC1", "EastGC7", "EastGC9", "EastGC10","EastGC11", "EastGC13", "EastGC2", "EastGC3", "EastGC4", "Private1", "Drain1"))

# Plot genetic lineage #
dapc.clus.p <- DAPC_2_probs %>% 
  ggplot(aes(SampleID, Probability, fill = factor(Cluster))) +
  geom_col(color = "gray", linewidth = 0.03, show.legend = F) +
  facet_grid(~WaterbodyName, 
             switch = "x", 
             scales = "free", 
             space = "free") +
  labs(x = "Waterbody", y = "Ancestry") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  #scale_fill_viridis_d(option = "plasma") +
  scale_fill_manual(values=clusCols) +
  ggtitle("DAPC cluster assignment")+
  theme(plot.title = element_text(size = 10)) +
  theme_minimal(base_size = 10) + 
  theme(panel.spacing.x = unit(0.001, "lines"),
        axis.text.x = element_blank(),
        panel.grid = element_blank()) + 
  #guides(fill = guide_legend(title = "Genetic Lineage")) +
  theme(strip.text.x = element_text(angle = 90, size = 12))


#set.seed(NULL)
```

&nbsp;

## Figure 3 - Combine PCA, bestK, overall DAPC
```{r Figure 3}
toprow <- ggarrange(pclrk.pca.p3, grp.K, labels = c("A", "B"), widths = c(1, 0.5))
botrow <- ggarrange(dapc.clus.p, labels = "C")

all_pca.dapc <- ggarrange(toprow, botrow, nrow = 2)

ggsave(all_pca.dapc, filename="figures/figure3_allPCA_bestk_dapc.png", width=40, height=30, dpi = 400, bg = "white")

```

&nbsp;

## Define waterbody colors
```{r}
# define colors
myCol <- brewer.pal(n = 10, name = 'Paired') 

popCols <- c("Apt1" = myCol[8], "Apt3" = myCol[8], "Apt2" = myCol[8], "WestGC1" = myCol[1], "WestGC2" = myCol[1], "WestGC3" = myCol[1], "WestGC6" = myCol[1], "Hotel5" = myCol[2], "Hotel1" = myCol[2], "Hotel2" = myCol[2], "Hotel3" = myCol[2], "Hotel4" = myCol[2], "EastGC11" = myCol[3],"EastGC10" = myCol[3],"EastGC7" = myCol[3], "EastGC9" = myCol[3], "EastGC1" = myCol[3], "EastGC13" = myCol[3], "EastGC2" = myCol[4], "EastGC3" = myCol[4], "EastGC4" = myCol[4], "Private1" = myCol[6], "Private2" = myCol[6], "Drain1" = myCol[9])
```

&nbsp;

## DAPC - Group 1
```{r}
g1.seed = 123
set.seed(g1.seed)

dat.genind.g1 <- poppr::popsub(dat.genind, sublist=c("Apt1", "Apt2", "Apt3")) #42 individuals; 2,672 loci

# dapc
dapc.g1 <- dapc(dat.genind.g1, var.contrib = TRUE, scale = FALSE, n.pca = 5,  n.da = nPop(dat.genind.g1) - 1)


# reformat to df and plot
dapc.g1.df <- cbind(as.data.frame(dapc.g1$ind.coord[,1:2]), dat.genind.g1@pop)
names(dapc.g1.df) <- c("A1", "A2", "pop")
centroids.g1 <- aggregate(cbind(A1, A2)~pop, dapc.g1.df, mean)
  

dapc.g1.p <- ggplot(dapc.g1.df, aes(x = A1, y = A2, color = pop)) +
  geom_point(alpha = 0.25, size = 3) +
  geom_point(data = centroids.g1, size = 5) +
  #geom_text_repel(data=centroids.g1, aes(label=pop), box.padding = .4, size=6) +
  geom_text_repel(data=centroids.g1, aes(label=pop, color="black"), box.padding = .4, size=6) +
  scale_color_manual(values = popCols) +
  ylab("Discriminant function 2") +
  xlab("Discriminant function 1") +
  stat_ellipse() +
  #ylim(-5,7) +
  #xlim(-1, 6) +
  theme_minimal(base_size = 16) +
  theme(legend.position="none") +
  guides(size = "none",
         shape = guide_legend(override.aes = list(size = 5)))

dapc.g1.p

set.seed(NULL)
```

&nbsp;

### Cluster analysis - Group 1
```{r}
#dat.genind.g1

# set seed
g1.seed = 111
set.seed(g1.seed)

# Run find.clusters
grp1.stats <- find.clusters(dat.genind.g1, max.n.clust=20, n.pca=200, choose.n.clust=FALSE)
g1.grp <- find.clusters(dat.genind.g1, max.n.clust=20, n.pca = 40, n.clust = 2)
# chose n.pca = 40, n.clusters= 2

g1.K <- grp1.stats$Kstat %>% as.data.frame() %>% mutate(K=1:20) %>% rename("BIC"=".") %>% filter(K < 7) %>%
ggplot(aes(x=K, y=BIC)) +
  geom_line(color="gray") +
  geom_point(shape=21, fill= "gray", size=3) +
  geom_vline(xintercept = length(g1.grp$size), linetype="dashed", color="black") +
  ggtitle("Group 1") +
  theme_minimal(base_size = 16)

g1.dapc1 <- dapc.genind(x = dat.genind.g1, pop = g1.grp$grp, n.pca = 40, n.da = 1) # n.pca=40, n.da=1

# Run optim a to find how many PCs to retain
g1.op <- optim.a.score(g1.dapc1)

# Rerun dapc with optim pcs
g1.dapc2 <- dapc.genind(x = dat.genind.g1, pop = g1.grp$grp, n.pca = 1, n.da = 1) # n.da=1

# Reformat dataframe
g1.DAPC_2_probs <- round(g1.dapc2$posterior, 2) %>%
  as_tibble(rownames = "SampleID") %>% 
  mutate(WaterbodyName = dat.genind.g1@pop) %>% 
  pivot_longer(cols = 2:3,   ## change depending on the nuROWer of groups
               names_to = "Cluster", 
               values_to = "Probability") %>% 
  mutate(WaterbodyName = fct_relevel(WaterbodyName, "Apt1", "Apt3", "Apt2"))


g1.col <- c(myCol[7], myCol[8], "gray")


g1.DAPC_2_probs <- g1.DAPC_2_probs %>% mutate(WaterbodyName = fct_relevel(WaterbodyName, "Apt1", "Apt3", "Apt2"))

# Plot genetic lineage #
g1.dapc.clus.p <- g1.DAPC_2_probs %>% 
  ggplot(aes(SampleID, Probability, fill = factor(Cluster))) +
  geom_col(color = "gray", linewidth = 0.03) +
  facet_grid(~WaterbodyName, 
             switch = "x", 
             scales = "free", 
             space = "free") +
  labs(x = "Waterbody", y = "Ancestry") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  #scale_fill_viridis_d(option = "plasma") +
  scale_fill_manual(values=g1.col) +
  ggtitle("ROW group")+
  theme(plot.title = element_text(size = 10)) +
  theme_minimal(base_size = 10) + 
  theme(panel.spacing.x = unit(0.001, "lines"),
        axis.text.x = element_blank(),
        panel.grid = element_blank()) + 
  guides(fill = guide_legend(title = "Genetic Lineage")) +
  theme(strip.text.x = element_text(angle = 90))

set.seed(NULL)
```

&nbsp;

## DAPC - Group 2
```{r}
dat.genind.Grp2 <- poppr::popsub(dat.genind, sublist=c("WestGC1", "WestGC2", "WestGC3", "WestGC6", "Hotel1", "Hotel2", "Hotel3", "Hotel4", "Hotel5")) # 333 individuals; 2,675 loci

# dapc
dapc.Grp2 <- dapc(dat.genind.Grp2, var.contrib = TRUE, scale = FALSE, n.pca = 60,  n.da = nPop(dat.genind.Grp2) - 1)


# reformat to df and plot
dapc.Grp2.df <- cbind(as.data.frame(dapc.Grp2$ind.coord[,1:7]), dat.genind.Grp2@pop)
names(dapc.Grp2.df) <- c("A1", "A2", "A3", "A4", "A5", "A6", "A7", "pop")
centroids.Grp2 <- aggregate(cbind(A1, A2)~pop, dapc.Grp2.df, mean)


dapc.Grp2.p <- ggplot(dapc.Grp2.df, aes(x = A1, y = A2, color = pop)) +
  geom_point(alpha = 0.25, size = 3) +
  geom_point(data = centroids.Grp2, size = 5) +
  #geom_text_repel(data=centroids.Grp2, aes(label=pop), box.padding = .4) +
  geom_text_repel(data=centroids.Grp2, aes(label=pop, color="black"), box.padding = 2.2, size=6) + #changed box pad from 0.4 
  scale_color_manual(values = popCols) +
  ylab("Discriminant function 2") +
  xlab("Discriminant function 1") +
  stat_ellipse() +
  #ylim(-5,7) +
  #xlim(-1, 6) +
  theme_minimal(base_size = 16) +
  theme(legend.position="none") +
  guides(size = "none",
         shape = guide_legend(override.aes = list(size = 5)))

dapc.Grp2.p
```

&nbsp;

### Cluster analysis - Group 2
```{r}
#dat.genind.Grp2

# set seed
Grp2.seed = 222
set.seed(Grp2.seed)

# Run find.clusters
grp2.stats <- find.clusters(dat.genind.Grp2, max.n.clust=20, n.pca=200, choose.n.clust=FALSE)
grp2.grp <- find.clusters(dat.genind.Grp2, max.n.clust=20, n.pca=200, n.clust=2)
# chose n.pca = 200, n.clusters= 2

grp2.K <- grp2.stats$Kstat %>% as.data.frame() %>% mutate(K=1:20) %>% rename("BIC"=".") %>% filter(K < 7) %>% 
ggplot(aes(x=K, y=BIC)) +
  geom_line(color="gray") +
  geom_point(shape=21, fill= "gray", size=3) +
  geom_vline(xintercept = length(grp2.grp$size), linetype="dashed", color="black") +
  ggtitle("Grp2 - MB+SBD") +
  theme_minimal(base_size = 16)

grp2.dapc1 <- dapc.genind(x = dat.genind.Grp2, pop = grp2.grp$grp, n.pca = 200, n.da = 1) # n.pca=200, n.da=1

# Run optim a to find how many PCs to retain
grp2.op <- optim.a.score(grp2.dapc1)

# Rerun dapc with optim pcs
grp2.dapc2 <- dapc.genind(x = dat.genind.Grp2, pop = grp2.grp$grp, n.pca = 1, n.da = 1) # n.da=1

# Reformat dataframe
grp2.DAPC_2_probs <- round(grp2.dapc2$posterior, 2) %>%
  as_tibble(rownames = "SampleID") %>% 
  mutate(WaterbodyName = dat.genind.Grp2@pop) %>% 
  pivot_longer(cols = 2:3,   ## change depending on the number of groups
               names_to = "Cluster", 
               values_to = "Probability") %>% 
  mutate(WaterbodyName = fct_relevel(WaterbodyName, "Hotel1", "Hotel2", "Hotel3", "Hotel4", "Hotel5", "WestGC1", "WestGC2", "WestGC3", "WestGC6"))



grp2.col <- c(myCol[1], myCol[2], "gray")


# Plot genetic lineage #
grp2.dapc.clus.p <- grp2.DAPC_2_probs %>% 
  ggplot(aes(SampleID, Probability, fill = factor(Cluster))) +
  geom_col(color = "gray", linewidth = 0.03) +
  facet_grid(~WaterbodyName, 
             switch = "x", 
             scales = "free", 
             space = "free") +
  labs(x = "Waterbody", y = "Ancestry") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  #scale_fill_viridis_d(option = "plasma") +
  scale_fill_manual(values=grp2.col) +
  ggtitle("Group2")+
  theme(plot.title = element_text(size = 10)) +
  theme_minimal(base_size = 10) + 
  theme(panel.spacing.x = unit(0.001, "lines"),
        axis.text.x = element_blank(),
        panel.grid = element_blank()) + 
  guides(fill = guide_legend(title = "Genetic Lineage")) +
  theme(strip.text.x = element_text(angle = 90))

set.seed(NULL)
```

&nbsp;

## DAPC - Group 3
```{r}
dat.genind.g3 <- poppr::popsub(dat.genind, sublist=c("EastGC11","EastGC10","EastGC7", "EastGC9","EastGC2","EastGC1", "EastGC3", "EastGC4", "EastGC13"))

# run dapc
dapc.g3 <- dapc(dat.genind.g3, var.contrib = TRUE, scale = FALSE, n.pca = 50,  n.da = nPop(dat.genind.g3) - 1)


# reformat to df and plot
dapc.g3.df <- cbind(as.data.frame(dapc.g3$ind.coord[,1:4]), dat.genind.g3@pop)
names(dapc.g3.df) <- c("A1", "A2", "A3", "A4", "pop")
centroids.FC <- aggregate(cbind(A1, A2)~pop, dapc.g3.df, mean)


dapc.g3.p <- ggplot(dapc.g3.df, aes(x = A1, y = A2, color = pop)) +
  geom_point(alpha = 0.25, size = 3) +
  geom_point(data = centroids.FC, size = 5) +
  #geom_text_repel(data=centroids.FC, aes(label=pop), box.padding = .4, size=6) +
  geom_text_repel(data=centroids.FC, aes(label=pop, color="black"), box.padding = .4, size=6) +
  scale_color_manual(values = popCols) +
  ylab("Discriminant function 2") +
  xlab("Discriminant function 1") +
  stat_ellipse() +
  #ylim(-5,7) +
  #xlim(-1, 6) +
  theme_minimal(base_size = 16) +
  theme(legend.position="none") +
  guides(size = "none",
         shape = guide_legend(override.aes = list(size = 5)))

dapc.g3.p

```

&nbsp;

### Cluster analysis - Group 3
```{r}
#dat.genind.FC

# set seed
g3.seed = 333
set.seed(g3.seed)

# Run find.clusters
grp3.stats <- find.clusters(dat.genind.g3, max.n.clust=20, n.pca=200, choose.n.clust=FALSE)
g3.grp <- find.clusters(dat.genind.g3, max.n.clust=20, n.pca=200, n.clust=3)
# chose n.pca = 200, n.clusters= 3

g3.K <- grp3.stats$Kstat %>% as.data.frame() %>% mutate(K=1:20) %>% rename("BIC"=".") %>% filter(K < 7) %>%
ggplot(aes(x=K, y=BIC)) +
  geom_line(color="gray") +
  geom_point(shape=21, fill= "gray", size=3) +
  geom_vline(xintercept = length(g3.grp$size), linetype="dashed", color="black") +
  ggtitle("Grp 3") +
  theme_minimal(base_size = 16)

g3.dapc1 <- dapc.genind(x = dat.genind.g3, pop = g3.grp$grp, n.pca = 200, n.da = 2) # n.pca=200, n.da=2

# Run optim a to find how many PCs to retain
g3.op <- optim.a.score(g3.dapc1)

# Rerun dapc with optim pcs
g3.dapc2 <- dapc.genind(x = dat.genind.g3, pop = g3.grp$grp, n.pca = 19, n.da = 2) # n.da=2

# Reformat dataframe
g3.DAPC_2_probs <- round(g3.dapc2$posterior, 2) %>%
  as_tibble(rownames = "SampleID") %>% 
  mutate(WaterbodyName = dat.genind.g3@pop) %>% 
  pivot_longer(cols = 2:4,   ## change depending on the number of groups
               names_to = "Cluster", 
               values_to = "Probability")

g3.col <- c(myCol[3], "gray", myCol[4])

# rename pops
g3.DAPC_2_probs <-  g3.DAPC_2_probs %>% mutate(WaterbodyName = fct_relevel(WaterbodyName, "EastGC1", "EastGC7", "EastGC9", "EastGC10","EastGC11", "EastGC13", "EastGC2", "EastGC3", "EastGC4"))


# Plot genetic lineage #
g3.dapc.clus.p <- g3.DAPC_2_probs %>% 
  ggplot(aes(SampleID, Probability, fill = factor(Cluster))) +
  geom_col(color = "gray", linewidth = 0.03) +
  facet_grid(~WaterbodyName, 
             switch = "x", 
             scales = "free", 
             space = "free") +
  labs(x = "Waterbody", y = "Ancestry") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = expansion(add = 1)) +
  #scale_fill_viridis_d(option = "plasma") +
  scale_fill_manual(values=g3.col) +
  ggtitle("g3 group")+
  theme(plot.title = element_text(size = 10)) +
  theme_minimal(base_size = 10) + 
  theme(panel.spacing.x = unit(0.001, "lines"),
        axis.text.x = element_blank(),
        panel.grid = element_blank()) + 
  guides(fill = guide_legend(title = "Genetic Lineage")) +
  theme(strip.text.x = element_text(angle = 90))

set.seed(NULL)
```

&nbsp;

#### Figure 4 - DAPC plots together 
```{r Figure 4}
# Plot best Ks for groups 1-3 
bestK.p2 <- ggarrange( g1.K + ggtitle("Group 1") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), text= element_text(size = 12)),
                       grp2.K + ggtitle("Group 2") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), text= element_text(size = 12)),
                       g3.K + ggtitle("Group 3") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), text= element_text(size = 12)), nrow=1)

bestK.p2.2 <- annotate_figure(bestK.p2, left = text_grob("BIC", rot = 90, size=16), bottom = text_grob("K", size=16))


# Combine DAPCs for groups 1-3
grp.dapcs <- ggarrange(dapc.g1.p + ggtitle("Group 1") + theme(axis.title.x = element_blank(), axis.title.y = element_blank(), title = element_text(size=12)),
                       dapc.Grp2.p + ggtitle("Group 2")+ theme(axis.title.x = element_blank(), axis.title.y = element_blank(), title = element_text(size=12)),
                       dapc.g3.p + ggtitle("Group 3")+ theme(axis.title.x = element_blank(), axis.title.y = element_blank(), title = element_text(size=12)),
                       nrow = 1 )

grp.dapcs.2 <- annotate_figure(grp.dapcs, left = text_grob("Discriminant function 2", rot = 90, size=18), bottom = text_grob("Discriminant function 1", size=18))


# combine cluster plots for groups 1-3
grp.dapcCluster <- ggarrange(g1.dapc.clus.p + ggtitle("Group 1") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), legend.position = "none"),
        grp2.dapc.clus.p + ggtitle("Group 2") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), legend.position = "none"),
        g3.dapc.clus.p + ggtitle("Group 3") + theme(axis.title.y = element_blank(), axis.title.x = element_blank(), legend.position = "none"),
          ncol = 1, align="h")

# combine cluster plots and dapc pca plots
dapcBigPlot <- ggarrange(grp.dapcs, grp.dapcCluster, nrow=2, labels = "AUTO")

# combine cluster plots and dapc pca plots AND best K - Figure 4
grp.dapcCluster.ann <- annotate_figure(grp.dapcCluster, left = text_grob("Ancestry", rot = 90, size=12))
dapcBigPlot2 <- ggarrange(grp.dapcs.2,bestK.p2.2, grp.dapcCluster.ann, nrow=3, labels = "AUTO", heights = c(1, 0.8, 0.8))

ggsave(dapcBigPlot2, filename="figures/figure4_hires.png", width=18, height=15, dpi = 600, bg = "white")

dapcBigPlot2
```